---
title: "Project 2: data preparation"
format: html
toc: TRUE
editor: visual
---

# Packages

Load the necessary packages.

```{r,warning=FALSE}

# Require will load the package if available and return FALSE if not available by adding the ! we will return TRUE if the package is not available, running the code in the if statement
if(!require(tidyverse)){
    # If package is not available, install it then load it again
    install.packages("tidyverse")
    library(tidyverse)
}

# Does same as above for other packages
if(!require(tidyr)){
    install.packages("tidyr")
    library(tidyr)
}

if(!require(readr)){
    install.packages("readr")
    library(readr)
}

if(!require(ggplot2)){
    install.packages("ggplot2")
    library(ggplot2)
}

if(!require(dplyr)){
    install.packages("dplyr")
    library(dplyr)
}

if(!require(janitor)){
    install.packages("janitor")
    library(janitor)
}  
if(!require(reshape2)){
    install.packages("reshape2")
    library(reshape2)
}

if(!require(shinycssloaders)){
    install.packages("shinycssloaders")
    library(shinycssloaders)
}

if(!require(bslib)){
    install.packages("bslib")
    library(bslib)
}

if(!require(DT)){
    install.packages("DT")
    library(DT)
}

if(!require(scales)){
    install.packages("scales")
    library(scales)
}

if(!require(forcats)){
    install.packages("forcats")
    library(forcats)
}


if(!require(stringr)){
    install.packages("stringr")
    library(stringr)
}


```

#Prepare your data
## Read in the dataset

I will be using the "Mobile Device Usage and User Behavior Dataset" from kaggle for this project.

```{r}
userBehavior <- read_csv("./data/user_behavior_dataset.csv")
userBehavior

# Janitor automatically fixes spaces, parentheses, slashes, and symbols
userBehavior <- userBehavior |>
  clean_names()
userBehavior
```

## Data Manipulation  

```{r}
#create a categorical variable based on a numeric variable age to create an age group and recode user_behavior_class variable to add descriptive label as the values are numeric 1-5.
userBehavior <- userBehavior |>
  mutate(age_group = case_when(
    age < 18 ~ "Under 18",
    age >= 18 & age < 25 ~ "18-24",
    age >= 25 & age < 35 ~ "25-34",
    age >= 35 & age < 45 ~ "35-44",
    age >= 45 & age < 60 ~ "45-59",
    age >= 60 ~ "60+",
    TRUE ~ NA_character_
  ),
  user_behavior_class_label = case_when(
    user_behavior_class == 1 ~ "Minimal User",
    user_behavior_class == 2 ~ "Light User",
    user_behavior_class == 3 ~ "Moderate User",
    user_behavior_class == 4 ~ "Heavy User",
    user_behavior_class == 5 ~ "Power User",
    TRUE ~ NA_character_))

# change datatype from character to factor, dbl to character for user ID.
userBehavior <- userBehavior |>
  mutate(user_id=as.character(user_id),
         operating_system = as.factor(operating_system),
         gender = as.factor(gender),
         user_behavior_class = as.factor(user_behavior_class)
         )

# to maintain my order and not let R order it alphabetically 
userBehavior$user_behavior_class_label <- factor(
  userBehavior$user_behavior_class_label,
  levels = c("Minimal User", "Light User", "Moderate User", "Heavy User", "Power User"),
  ordered = TRUE
)
userBehavior

# save my derived data
saveRDS(userBehavior, file = "data/userbehavior_der.rds")
write.csv(userBehavior,"data/userbehavior_der.csv",row.names = FALSE)
```

## Explore the dataset 
### One-Way contingency tables
```{r}
#these one way contingency tables show the gender, operating system and user behavior distribution in the dataset.
tab_gender <- table(userBehavior$gender)
tab_gender

tab_operating_system <- table(userBehavior$operating_system)
tab_operating_system


tab_behavior_class <- table(userBehavior$user_behavior_class_label)
tab_behavior_class
```
### Two-Way contingency tables
```{r}
# Two-way tables (joint frequency)
table_behavior_os <- table(userBehavior$user_behavior_class_label, userBehavior$operating_system)
table_behavior_os

table_behavior_gender <- table(userBehavior$gender , userBehavior$user_behavior_class_label )
table_behavior_gender

table_os_gender <- table(userBehavior$operating_system, userBehavior$gender)
table_os_gender
prop.table(table_os_gender, margin = 2)  # proportions within gender

```
### Numerical summaries for quantitative variables by categorical groups

```{r}
# we can use the across() function and summarize all numeric variables at once 
userBehavior |>
  summarize(across(where(is.numeric), 
                   list("mean" = mean, "median" = median,"Std"=sd) , 
                   .names = "{.fn}_{.col}"))

# This is summary of quantitative variables by user behavior classes (categorical). 
userBehavior |>
  group_by(user_behavior_class_label) |>
  summarise(
    mean_app_usage = mean(app_usage_time_min_day, na.rm = TRUE),
    median_app_usage = median(app_usage_time_min_day, na.rm = TRUE),
    sd_app_usage = sd(app_usage_time_min_day, na.rm = TRUE),
    mean_data_usage = mean(data_usage_mb_day, na.rm = TRUE),
    median_data_usage = median(data_usage_mb_day, na.rm = TRUE),
    sd_data_usage = sd(data_usage_mb_day, na.rm = TRUE)
    
  )

```
### Graphical Summaries

#### Plot 1: Distribution of user behavior class (Univariate Bar Plot)


```{r}
ggplot(userBehavior|> drop_na(user_behavior_class_label), aes(x = user_behavior_class_label, fill = user_behavior_class_label)) +
  geom_bar() +
  scale_fill_manual(
    values = c(
      "Minimal User" = "#00B050",   # green
      "Light User"   = "#92D050",   # light green
      "Moderate User"= "#FFC000",   # yellow
      "Heavy User"   = "#FF7043",   # orange
      "Power User"   = "#C00000"    # red
    ),
     name = "User Behavior Classes"
  ) +
  labs(title = "Distribution of User Behavior Classes",
       x = "User Behavior Class", y = "Count") +
  theme_minimal()
```


#### Plot 2: App Usage Time by Behavior Class (Boxplot)

```{r}
ggplot(userBehavior |> drop_na(user_behavior_class, app_usage_time_min_day), 
       aes(x = user_behavior_class, y = app_usage_time_min_day, fill = user_behavior_class_label)) +
  geom_boxplot() +
   scale_fill_manual(
    values = c(
      "Minimal User" = "#00B050",   # green
      "Light User"   = "#92D050",   # light green
      "Moderate User"= "#FFC000",   # yellow
      "Heavy User"   = "#FF7043",   # orange
      "Power User"   = "#C00000"    # red
    ),
     name = "User Behavior Classes"
   ) +
  labs(title = "App Usage Time by User Behavior Class",
       x = "User Behavior Class", y = "App Usage (min/day)") +
  theme_minimal()

```

#### Plot 3: App usage by Operating system (histogram)
l
```{r}
g <- ggplot(userBehavior |> drop_na(app_usage_time_min_day, operating_system) ,
 aes(x = app_usage_time_min_day,  fill = operating_system))
g + geom_histogram(position = "identity", bins = 35,alpha = 0.5  ) +
 
  labs(
    title = "Distribution of App Usage by Operating System",
    X = "App Usage (min/day)",
    y = "count",
    fill="operating_system"
  ) +
  theme_minimal()
```



#### Plot 4: Distribution of operating system by gender  (Bar chart)

```{r}
g <- ggplot(data = userBehavior |> drop_na(gender, operating_system), aes(x = gender, fill = operating_system))
g + geom_bar(position = "dodge")+
  labs(title = "Operating System Type by Gender", 
       x = "Gender")+
 scale_fill_discrete("Operating System")
```


#### Plot 5: Data Usage by Behavior Class and Gender (Multivariate Boxplot with Grouping)

```{r}
# Multivariate (Behavior class X Gender)
ggplot(userBehavior, aes(x = user_behavior_class, y = data_usage_mb_day, fill = gender)) +
  geom_boxplot(position = position_dodge(width = 0.8)) +
  labs(title = "Data Usage by Behavior Class and Gender",
       x = "User Behavior Class", y = "Data Usage (MB/day)", fill = "Gender") +
  theme_minimal()

```

#### Plot 6: App Usage vs Data Usage colored by Operating System (Scatterplot)

```{r}
ggplot(userBehavior, aes(x = app_usage_time_min_day, y = data_usage_mb_day, color = operating_system)) +
  geom_point(alpha = 0.7) +
  geom_smooth(method = "lm", se = FALSE, color = "black") +
  labs(title = "Relationship Between App Usage and Data Usage by OS",
       x = "App Usage (min/day)", y = "Data Usage (MB/day)", color = "OS") +
  theme_minimal()

```

#### Plot 7: App Usage vs Data Usage colored by Operating System (Scatterplot)

```{r}
ggplot(userBehavior, aes(x = number_of_apps_installed,
                 y = screen_on_time_hours_day,
                 color = age_group,
                 size = age_group,
                 )) +
  geom_point(alpha = 0.7, size = 3)  +
 scale_color_brewer(palette = "Spectral")  +
  labs(
    title = "Screen Time vs Number of Apps by Age Group",
    x = "Number of Apps Installed",
    y = "Screen On Time (hours/day)",
    color = "Age Group",
    
  ) +
  theme_minimal()

```
#### Plot 8: Screen time by gender (Boxplots with Points)
l
```{r}
g <- ggplot(userBehavior |> drop_na(screen_on_time_hours_day, gender) ,
 aes(x = gender, y = screen_on_time_hours_day, fill = gender))
g + geom_boxplot() +
 geom_jitter(width = 0.2, alpha = 0.3)+
  labs(
    title = "Screen Time  by Gender",
    y = "Screen On Time (hours/day)",
    x = "Gender"
  ) +
  theme_minimal()
```

#### Plot 9: Faceted Boxplots for Multiple Metrics

```{r}
userBehavior |>
  select(user_behavior_class, app_usage_time_min_day, screen_on_time_hours_day, battery_drain_m_ah_day) |>
  pivot_longer(-user_behavior_class, names_to = "metric", values_to = "value") |>
  ggplot(aes(x = user_behavior_class, y = value, fill = user_behavior_class)) +
  geom_boxplot() +
  facet_wrap(~ metric, scales = "free_y") +
  labs(title = "Device Metrics by User Behavior Class",
       x = "User Behavior Class", y = "Value") +
  theme_minimal() +
  theme(legend.position = "none")

```
#### Plot 10: screen time vs app usage by age group (faceted Scatter plot)
Insight: Reveals which devices or OS are more power-efficient under high usage.

```{r}
ggplot(userBehavior, aes(x = battery_drain_m_ah_day,
                 y = screen_on_time_hours_day,
                 color = device_model)) +
  geom_point(alpha = 0.7, size = 2.5) +
  facet_wrap(~operating_system) +
  labs(
    title = "Battery drain vs Screen Time by Operating System",
    x = "Battery Drain (Min/day)",
    y = "Screen On Time (hours/day)"
  ) +
  theme_minimal()

```

#### Plot 11: Heatmap of Correlations (Not covered in class)

```{r}
# Select numeric variables
# Select numeric variables
num_vars <- userBehavior %>%
  select(app_usage_time_min_day, screen_on_time_hours_day,
         battery_drain_m_ah_day, number_of_apps_installed,
         data_usage_mb_day, age)

# Compute correlation matrix
cor_matrix <- round(cor(num_vars, use = "pairwise.complete.obs"), 2)

# Melt for ggplot
cor_long <- melt(cor_matrix)

# Plot heatmap (0 to 1 scale)
ggplot(cor_long, aes(Var1, Var2, fill = value)) +
  geom_tile(color = "white") +
  geom_text(aes(label = round(value, 2)), color = "black", size = 3) +
  scale_fill_gradientn(
    colours = c("white", "yellow", "orange", "red", "darkred"),
    limits  = c(0, 1),
    name    = "Correlation"
  ) +
  labs(title = "Heatmap of Positive Correlations (0–1)",
       x = NULL, y = NULL) +
  theme_minimal(base_size = 12) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```



